The influence of Gaia-Sausage-Enceladus-like merger events on the chemo-dynamics of galaxy discs

### 2023-01-30 src/tex/ms.tex
[showyourwork] overleaf sync

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[Early massive accretion events in a MW-mass galaxy]{The influence of massive early merger events on the chemo-dynamics of galaxy discs}

@@

@@

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events. Outside of a radius of $\sim5$ kpc the merging satellites contribute cold gas up to a factor of 4 in surface density.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

\begin{figure*}
    \script{feh_gradient.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/feh_gradient.pdf}
        \caption{
            Metallicity gradient of the cold gas disk during the merger events. We find that the contribution of additional metal poor gas lowers the metallicity in the outskirts of the disk effectively steepening the metallicity gradient by $\sim 0.1$ dex.
        }
        \label{fig:feh_grad}
    \end{centering}
\end{figure*}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[Early massive accretion events in MW-mass galaxies]{The influence of Gaia-Sausage-Enceladus-like merger events on the chemo-dynamics of galaxy discs}
%The influence of massive early merger events on the chemo-dynamics of galaxy discs

@@

Our Milky Way (MW) Galaxy is the best-studied galaxy in the
universe and provides the most stringent constraints of galaxy formation
models \citep[e.g.,][]{Guedes11, Wetzel16, Grand17, Buck21}. Therefore, a thorough understanding of the components of the Galaxy, their dynamical evolution and formation channel is of great interest.

The Galaxy's main body, the rotationally-supported stellar disk harbors multiple components or populations when dissected in the space of chemistry, kinematics, spatial extent, and age \citep[e.g.,][]{Gilmore83, Norris85, Chiba00, Nissen10, Bovy12, Haywood13}; see \citet{Rix13} and \citet{Bland-Hawthorn16} for recent reviews. %There exist a dichotomy in both spatial structure and stellar chemistry. Spatially, MW's disk can be separated into a thin and a thick disk with different scale heights. Chemically, a clear bimodality in the abundance of $\alpha$ elements relative to iron ([$\alpha$/Fe]) has been found. In general, the high$-\alpha$ population has older ages and a larger scale-height than the low$-\alpha$ population. However, the mapping between the high$-\alpha$ (low$-\alpha$) and the thick (thin) disks is not perfect.

Thanks to the great improvements in the quality and volume of astrometric datasets provided by the {\it Gaia} mission in combination with chemical abundances and radial velocities from large spectroscopic surveys, (e.g. \textsc{apogee}; \textsc{galah}; \textsc{h3}; \textsc{LAMOST}; \citealt{Majewski2017}, \citealt{Martell2017}, \citealt{Conroy2019}, \citealt{Zhao_2012}, respectively) we are now able to identify and link coherent chemodynamical structures in the solar neighbourhood to ancient merger events several billion years back in MW's history (e.g. \citealt{Helmi2020} for a review). By now a large number of chemocynamical clusters associated with accreted systems have been identified in the high-dimensional space of orbital parameters and stellar abundances \citep[e.g.,][]{Belokurov18, Helmi18, Myeong19a, Naidu20a, Horta21, Myeong2022}.

Most notably, there exists an excess of stars on radial orbits in the local stellar halo around the Sun which is often referred to as {\it Gaia}-Sausage-Enceladus (GSE; \citealt{Belokurov2018, Helmi2018}; see also \citealt{Nissen2010, Koppelman2018, Haywood2018}). This stellar overdensity has been identified as the observable signature of a massive ($M_{\star} \simeq 10^{9}$ $\rm{M_{\odot}}$) early ($8-11$ Gyr ago; e.g. \citealt{Vin2019, Bel_2020, Naidu21, Xiang22}) accretion event. GSE is by far the most significant merger in MW's history and contributed approximately two-thirds of MW's stellar halo stars on highly-eccentric orbits \citep[e.g.][]{Mackereth+Bovy20}. GSE stars inhabit a sausage-like distribution in the radial-azimuthal velocity distribution \citep{Brook2003, Belokurov2018} and appear to be more metal-poor and less $\alpha$-enhanced than the redder halo counterpart \citep{Haywood_2018a, Helmi_2018}. The redder sequence of the Galactic halo is thought to be the result of proto-galactic disc stars being dynamically ejected into the halo during the GSE merger and has been associated with an event termed {\it the Splash} by \cite{Belokurov2020, Bonaca2020}, initially also found by other earlier studies \citep{Bonaca2017, Haywood2018, DiMatteo2019, Gallart2019}.

The existence of the GSE merger and its strong impact on the structure of the proto-MW demands a thorough study of the processes that shape MW's chemo-dynamical evolution during such a merger in order to interpret the observational data at hand. In order to connect present day observables to Gyr old events we need tools like cosmological simulations that model galactic mass growth, mergers and subsequent star formation plus chemical enrichment self-consistently. Here, enough resolution to follow the internal disk dynamics and suppress spurious heating ($N \gtrsim 10^6$; e.g. \citealt{Sellwood2013,Ludlow2019DiskHeating, Ludlow2021}) as well as resolving the multi-phase, dense structure of the interstellar medium (ISM) combined with a proper model for chemical enrichment \citep[e.g.][]{Buck2021} is key to accurately capture their birth kinematics and the subsequent dynamical evolution of the disk (e.g. \citealt{House2011, Bird2013}).

Thanks to recent advances in computing power and progress in numerical methods modern cosmological zoom simulations now meet the aforementioned requirements and are able to model individual MW-mass galaxies by sampling varying environments and formation scenarios (e.g. \citealt{Sawala2016, Grand2017, Buck2020, Font2020, Applebaum2021, Agertz2020Vintergatan, Bird2021, Khoperskov2022-1InSitu, Wetzel2022}). This enables us to reconstruct our Galaxy's formation history by linking the occurrence of chemodynamical patterns at $z=0$ to specific events in the Galaxy's (e.g. \citealt{Bignone2019, Mackereth2019, Fattahi2019, Grand2020, Elias2020, Dillamore2022, Khoperskov2022-2Accreted, Khoperskov2022-3Metallicity, Pagnini2022, Rey2022}).

While GSE's strong impact on MW's stellar halo is by now well established, more recently studies have further investigated its defining influence on the formation of MW's thin and thick disk \citep[e.g.]{Grand2020,Ciuca2022,Orkney2022}. These studies suggest that the GSE merger strongly shaped the formation of MW's thick disk agreeing with thick disk formation scenarios from early simulation results by \cite{Brook_2004, Brook_2006}. The gas-rich merger has a two-fold effect on MW's stellar disk: (i) Firstly, it heats part of the existing proto-disc stars dynamically ejecting them into the halo and creating the Splash. (ii) Secondly, the merger provides fresh gas to the central galactic regions triggering a starburst that eventually forms the younger thick disc. After the merge, the thin disc then forms from the accretion of metal-poor gas in an inside-out, upside-down fashion \citep[e.g.,][]{Bird2013, Grand2018, Buck2020a}. \citet{Ciuca_2021} qualitatively confirmed this picture using APOGEE DR14 data and selecting stellar populations around the solar radius.

Interestingly, when inferring the birth radii of MW disk stars, \citet{Lu2021} found a steepening of the metallicity gradient at the time of the GSE merger using data from LAMOST combined with {\it Gaia} eDR3. Here we use the NIHAO-UHD suite \citep{Buck2020} of cosmological hydrodynamical siumulations of MW-mass galaxies in order to investigate the influence of early merger events on the chemodynamics of the stellar disk. By using 4 different simulations from the NIHAO-UHD suite we investigate independently sampled environments that follow 4 different formation scenarios. Especially we focus on the frequency of massive early mergers and their detailed impact on the formation of the stellar disk. This paper is tructures as follows: In \S 2 we describe the simulations on which we base our analysis, followed by a presentation of our results in \S 3. In \S 4 we discuss and conclude our findings.

@@

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events. Outside of a radius of $\sim5-10$ kpc the merging satellites contribute significant amounts of cold gas up to 200\%.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

\begin{figure}
    \script{feh_gradient.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/feh_gradient.pdf}
        \caption{
            Metallicity gradient of the cold gas disk during the merger events. We find that the contribution of additional metal poor gas lowers the metallicity in the outskirts of the disk effectively steepening the metallicity gradient by $\sim 0.1$ dex.
        }
        \label{fig:feh_grad}
    \end{centering}
\end{figure}
